\textbf{The Illusion Game: A Novel Experimental Paradigm Provides Evidence in Favour of a General Factor of Visual Illusion Sensitivity and Personality Correlates}

### 2022-09-23 submission/manuscript.Rmd
update

  - id            : "1"
    institution   : "School of Social Sciences, Nanyang Technological University, Singapore"
  - id            : "2"
    institution   : "LKC Medicine, Nanyang Technological University, Singapore"
  - id            : "3"
    institution   : "National Institute of Education, Singapore"
  - id            : "4"
    institution   : "Centre for Research and Development in Learning, Nanyang Technological University, Singapore"
authornote: |
  Correspondence concerning this article should be addressed to Dominique Makowski, HSS 04-18, 48 Nanyang Avenue, Singapore (dom.makowski@gmail.com).
abstract: |
  Visual illusions highlight how the brain uses contextual and prior information to inform our perception of reality. Unfortunately, illusion research has been hampered by the difficulty of adapting these stimuli to experimental settings. In this study, we used the novel parametric framework for visual illusions to generate 10 different classic illusions (Delboeuf, Ebbinghaus, Rod and Frame, Vertical-Horizontal, Zöllner, White, Müller-Lyer, Ponzo, Poggendorff, Contrast) varying in strength, embedded in a perceptual discrimination task. We tested the objective effect of the illusions on errors and reaction times, and extracted participant-level performance scores (n=250). Our results provide evidence in favour of a general factor (labelled Factor *i*) underlying the sensitivity to different illusions. Moreover, we report a positive relationship between illusion sensitivity and personality traits such as Agreeableness, Honesty-Humility, and negative relationships with Psychoticism, Antagonism, Disinhibition, and Negative Affect.
keywords          : "visual illusions, illusion game, Pyllusion, personality, general factor"
wordcount         : "3876"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

One area of contention concerns the presence of a common mechanism underlying the effect of different illusions [@hamburger2016; @cretenoud2020illusions]. While early research has suggested a common factor of illusion sensitivity indexed by overall vision proficiency [@halpern1999interindividual; @thurstone1944factorial], recent empirical studies at most observed weak correlations between resistance to distinct illusions [@grzeczkowski2017; @grzeczkowski2018]. The existence of dispositional correlates of illusion sensitivity has also been controversial, related to evidence suggesting a lower illusion sensitivity in in patients with schizophrenia and autism [@notredame2014; @park2022; @grzeczkowski2018; @gori2016; @razeghi2022], as well as individuals with stronger aggression and narcissism traits [@zhang2017; @konrath2009seeing].

@@

The inter-individual variability in the effect of illusion strength and its interaction with task difficulty was extracted from the models and used as participant-level scores. We then explored the relationship of these indices across different illusions using exploratory factor analysis (EFA) and structural equation modelling (SEM), and tested the existence of a general of illusion sensitivity (Factor *i*).

@@

For RTs, the GAMs suggested a consistent non-linear relationship between RT and illusion strength: as the illusion strength increases beyond a certain threshold, the participants responded faster. While this is is not surprising (strong illusions are likely so effective in biasing perception that it is "easier", i.e., faster, to make the wrong decision), the linear models were not designed to capture this - likely quadratic - pattern and hence are not good representatives of the underlying dynamics. As such, we decided not to use them for the individual scores analysis. <!-- Additionally, it would have increased the parameter space from 20 to 40 variables, which was deemed as statistically unreasonable given our sample size). -->

@@

The Bayesian correlation analysis (with narrow priors centered around a null effect) between the illusion scores and contextual variables (screen size and refresh rate) provided weak evidence in favor of an absence of effect, with the exception of the two contrast-based illusions (**Figure 3**). Anecdotal ($BF_{10} = 2.05$) and moderate evidence ($BF_{10} = 4.11$) was found for a negative correlation between screen size and the sensitivity to the White and the Contrast illusion, respectively. To test whether this result could be an artifact related to the highly skewed screen size distribution (caused by very few participants with extreme screen sizes), we re-ran a robust correlation (with rank-transformed values), which provided even stronger evidence in favor of the effect existence ($BF_{10} = 28.19$, $BF_{10} = 4.31$ for White and Contrast, respectively).

@@

The Bayesian t-tests on the effect of sex suggested anecdotal to moderate evidence in favour of the null effect for all scores, with the exception of the sensitivity to the Zöllner illusion, which was higher in males as compared to females ($\Delta=-0.37$, 95% CI [-0.62, -0.13], $BF_{10} = 12.74$). We fitted Bayesian linear models with the education level entered as a monotonic predictor [appropriate for ordinal variables, @burkner2020modelling], which yielded no significant effects. For age, we fitted two types of models for each score, one general additive models (GAM) and a 2nd order polynomial model. These consistently suggested a significant positive linear relationship between age and Factor *i* ($pd=100\%$), as well as the sensitivity to Müller-Lyer ($pd=100\%$), Vertical-Horizontal ($pd=100\%$), Zöllner ($pd=100\%$) and Ebbinghaus ($pd=99\%$) illusions.

>>>>

  - id            : "1"
    institution   : "School of Social Sciences, Nanyang Technological University, Singapore"
  - id            : "2"
    institution   : "LKC Medicine, Nanyang Technological University, Singapore"
  - id            : "3"
    institution   : "National Institute of Education, Singapore"
  - id            : "4"
    institution   : "Centre for Research and Development in Learning, Nanyang Technological University, Singapore"
authornote: |
  Correspondence concerning this article should be addressed to Dominique Makowski, HSS 04-18, 48 Nanyang Avenue, Singapore (dom.makowski@gmail.com).
abstract: |
  Visual illusions highlight how the brain uses contextual and prior information to inform our perception of reality. Unfortunately, illusion research has been hampered by the difficulty of adapting these stimuli to experimental settings. In this study, we used the novel parametric framework for visual illusions to generate 10 different classic illusions (Delboeuf, Ebbinghaus, Rod and Frame, Vertical-Horizontal, Zöllner, White, Müller-Lyer, Ponzo, Poggendorff, Contrast) varying in strength, embedded in a perceptual discrimination task. We tested the objective effect of the illusions on errors and reaction times, and extracted participant-level performance scores (n=250). Our results provide evidence in favour of a general factor (labelled Factor *i*) underlying the sensitivity to different illusions. Moreover, we report a positive relationship between illusion sensitivity and personality traits such as Agreeableness, Honesty-Humility, and negative relationships with Psychoticism, Antagonism, Disinhibition, and Negative Affect.
keywords          : "visual illusions, illusion game, Pyllusion, personality, general factor"
wordcount         : "3873"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

One area of contention concerns the presence of a common mechanism underlying the effect of different illusions [@hamburger2016; @cretenoud2020illusions]. While early research has suggested a common factor of illusion sensitivity indexed by overall vision proficiency [@halpern1999interindividual; @thurstone1944factorial], recent empirical studies observed at nost weak correlations between resistance to distinct illusions [@grzeczkowski2017; @grzeczkowski2018]. The existence of dispositional correlates of illusion sensitivity has also been controversial, with evidence suggesting a lower illusion sensitivity in patients with schizophrenia and autism [@notredame2014; @park2022; @grzeczkowski2018; @gori2016; @razeghi2022], as well as individuals with stronger aggression and narcissism traits [@zhang2017; @konrath2009seeing].

@@

The inter-individual variability in the effect of illusion strength and its interaction with task difficulty was extracted from the models and used as participant-level scores. We then explored the relationship of these indices across different illusions using exploratory factor analysis (EFA) and structural equation modelling (SEM), and tested the existence of a general factor of illusion sensitivity (Factor *i*).

@@

For RTs, the GAMs suggested a consistent non-linear relationship between RT and illusion strength: as the illusion strength increases beyond a certain threshold, the participants responded faster. While this is not surprising (strong illusions are likely so effective in biasing perception that it is "easier", i.e., faster, to make the wrong decision), the linear models were not designed to capture this - likely quadratic - pattern and hence are not good representatives of the underlying dynamics. As such, we decided not to use them for the individual scores analysis. <!-- Additionally, it would have increased the parameter space from 20 to 40 variables, which was deemed as statistically unreasonable given our sample size). -->

@@

The Bayesian correlation analysis (with narrow priors centered around a null effect) between the illusion scores and contextual variables (screen size and refresh rate) provided weak evidence in favor of an absence of effect, with the exception of the two contrast-based illusions. Anecdotal ($BF_{10} = 2.05$) and moderate evidence ($BF_{10} = 4.11$) was found for a negative correlation between screen size and the sensitivity to the White and the Contrast illusion, respectively. To test whether this result could be an artifact related to the highly skewed screen size distribution (caused by very few participants with extreme screen sizes), we re-ran a robust correlation (with rank-transformed values), which provided even stronger evidence in favor of the effect existence ($BF_{10} = 28.19$, $BF_{10} = 4.31$ for White and Contrast, respectively).

The Bayesian t-tests on the effect of sex suggested anecdotal to moderate evidence in favour of the null effect for all scores, with the exception of the sensitivity to the Zöllner illusion, which was higher in males as compared to females ($\Delta=-0.37$, 95% CI [-0.62, -0.13], $BF_{10} = 12.74$). We fitted Bayesian linear models with the education level entered as a monotonic predictor [appropriate for ordinal variables, @burkner2020modelling], which yielded no significant effects. For age, we fitted two types of models for each score, one general additive models (GAM) and a 2nd order polynomial model. These consistently suggested a significant positive linear relationship between age and Factor *i* ($pd=100\%$), as well as the sensitivity to Müller-Lyer ($pd=100\%$), Vertical-Horizontal ($pd=100\%$), Zöllner ($pd=100\%$) and Ebbinghaus ($pd=99\%$) illusions (**Figure 3**).

@@

### 2022-09-23 submission/manuscript.tex
update

One area of contention concerns the presence of a common mechanism underlying the effect of different illusions (Cretenoud et al., 2020; Hamburger, 2016). While early research has suggested a common factor of illusion sensitivity indexed by overall vision proficiency (Halpern et al., 1999; Thurstone, 1944), recent empirical studies at most observed weak correlations between resistance to distinct illusions (Grzeczkowski et al., 2017, 2018). The existence of dispositional correlates of illusion sensitivity has also been controversial, related to evidence suggesting a lower illusion sensitivity in in patients with schizophrenia and autism (Gori et al., 2016; Grzeczkowski et al., 2018; Notredame et al., 2014; Park et al., 2022; Razeghi et al., 2022), as well as individuals with stronger aggression and narcissism traits (Konrath et al., 2009; Zhang et al., 2017).

@@

The inter-individual variability in the effect of illusion strength and its interaction with task difficulty was extracted from the models and used as participant-level scores. We then explored the relationship of these indices across different illusions using exploratory factor analysis (EFA) and structural equation modelling (SEM), and tested the existence of a general of illusion sensitivity (Factor \emph{i}).

@@

For RTs, the GAMs suggested a consistent non-linear relationship between RT and illusion strength: as the illusion strength increases beyond a certain threshold, the participants responded faster. While this is is not surprising (strong illusions are likely so effective in biasing perception that it is ``easier'', i.e., faster, to make the wrong decision), the linear models were not designed to capture this - likely quadratic - pattern and hence are not good representatives of the underlying dynamics. As such, we decided not to use them for the individual scores analysis.

@@

The Bayesian correlation analysis (with narrow priors centered around a null effect) between the illusion scores and contextual variables (screen size and refresh rate) provided weak evidence in favor of an absence of effect, with the exception of the two contrast-based illusions (\textbf{Figure 3}). Anecdotal (\(BF_{10} = 2.05\)) and moderate evidence (\(BF_{10} = 4.11\)) was found for a negative correlation between screen size and the sensitivity to the White and the Contrast illusion, respectively. To test whether this result could be an artifact related to the highly skewed screen size distribution (caused by very few participants with extreme screen sizes), we re-ran a robust correlation (with rank-transformed values), which provided even stronger evidence in favor of the effect existence (\(BF_{10} = 28.19\), \(BF_{10} = 4.31\) for White and Contrast, respectively).

@@

The Bayesian t-tests on the effect of sex suggested anecdotal to moderate evidence in favour of the null effect for all scores, with the exception of the sensitivity to the Zöllner illusion, which was higher in males as compared to females (\(\Delta=-0.37\), 95\% CI {[}-0.62, -0.13{]}, \(BF_{10} = 12.74\)). We fitted Bayesian linear models with the education level entered as a monotonic predictor (appropriate for ordinal variables, Bürkner \& Charpentier, 2020), which yielded no significant effects. For age, we fitted two types of models for each score, one general additive models (GAM) and a 2nd order polynomial model. These consistently suggested a significant positive linear relationship between age and Factor \emph{i} (\(pd=100\%\)), as well as the sensitivity to Müller-Lyer (\(pd=100\%\)), Vertical-Horizontal (\(pd=100\%\)), Zöllner (\(pd=100\%\)) and Ebbinghaus (\(pd=99\%\)) illusions.

>>>>

One area of contention concerns the presence of a common mechanism underlying the effect of different illusions (Cretenoud et al., 2020; Hamburger, 2016). While early research has suggested a common factor of illusion sensitivity indexed by overall vision proficiency (Halpern et al., 1999; Thurstone, 1944), recent empirical studies observed at nost weak correlations between resistance to distinct illusions (Grzeczkowski et al., 2017, 2018). The existence of dispositional correlates of illusion sensitivity has also been controversial, with evidence suggesting a lower illusion sensitivity in patients with schizophrenia and autism (Gori et al., 2016; Grzeczkowski et al., 2018; Notredame et al., 2014; Park et al., 2022; Razeghi et al., 2022), as well as individuals with stronger aggression and narcissism traits (Konrath et al., 2009; Zhang et al., 2017).

@@

The inter-individual variability in the effect of illusion strength and its interaction with task difficulty was extracted from the models and used as participant-level scores. We then explored the relationship of these indices across different illusions using exploratory factor analysis (EFA) and structural equation modelling (SEM), and tested the existence of a general factor of illusion sensitivity (Factor \emph{i}).

@@

For RTs, the GAMs suggested a consistent non-linear relationship between RT and illusion strength: as the illusion strength increases beyond a certain threshold, the participants responded faster. While this is not surprising (strong illusions are likely so effective in biasing perception that it is ``easier'', i.e., faster, to make the wrong decision), the linear models were not designed to capture this - likely quadratic - pattern and hence are not good representatives of the underlying dynamics. As such, we decided not to use them for the individual scores analysis.

@@

The Bayesian correlation analysis (with narrow priors centered around a null effect) between the illusion scores and contextual variables (screen size and refresh rate) provided weak evidence in favor of an absence of effect, with the exception of the two contrast-based illusions. Anecdotal (\(BF_{10} = 2.05\)) and moderate evidence (\(BF_{10} = 4.11\)) was found for a negative correlation between screen size and the sensitivity to the White and the Contrast illusion, respectively. To test whether this result could be an artifact related to the highly skewed screen size distribution (caused by very few participants with extreme screen sizes), we re-ran a robust correlation (with rank-transformed values), which provided even stronger evidence in favor of the effect existence (\(BF_{10} = 28.19\), \(BF_{10} = 4.31\) for White and Contrast, respectively).

The Bayesian t-tests on the effect of sex suggested anecdotal to moderate evidence in favour of the null effect for all scores, with the exception of the sensitivity to the Zöllner illusion, which was higher in males as compared to females (\(\Delta=-0.37\), 95\% CI {[}-0.62, -0.13{]}, \(BF_{10} = 12.74\)). We fitted Bayesian linear models with the education level entered as a monotonic predictor (appropriate for ordinal variables, Bürkner \& Charpentier, 2020), which yielded no significant effects. For age, we fitted two types of models for each score, one general additive models (GAM) and a 2nd order polynomial model. These consistently suggested a significant positive linear relationship between age and Factor \emph{i} (\(pd=100\%\)), as well as the sensitivity to Müller-Lyer (\(pd=100\%\)), Vertical-Horizontal (\(pd=100\%\)), Zöllner (\(pd=100\%\)) and Ebbinghaus (\(pd=99\%\)) illusions (\textbf{Figure 3}).

@@
